Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-21 paper.Rmd
next to last on Monday

---
title: "Reassessing the role of theory and machine learning in forecasting civil conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory forecasting is more accurate than has previously been possible. Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. Actually, it does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the original unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.\footnote{We set aside for here the characteristics of theoretical models, accepting the B\&S denotation.}  They also claim that the integration of structure with process is better than alternaties over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

1. Split the training data into training (2001 - 2007) and test (2008 - 2015) sets.
2. Estimate the escalation and other competing models. 
3. Create out-of-sample (OOS) predictions from each model using the test set. 
4. Calculate AUC-ROC measures for each set of OOS predictions.

@@

The corresponding results for each question are shown in B&S Tables 1 and 2, which we examine further below. We accurately replicate their Tables 1 and 2, with very tiny differences. The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

@@

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model is not the best, and it actually performs worse than the atheoretical, garbage can model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey.} (Both of the ambiguous cases include Blair as coauthor.) The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

@@

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,beger:etal:2017]. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually just creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking just another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

We now turn to an examination that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2). We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S, and may an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example is found in The Political Logic of Survival [@bdm:etal:2005]. For the purposes of discussion we call this strong theory.

@@

At this level of detail the theoretical model is more about choosing the correct variables, and this is the contrast between theoretical prediction models and a-theoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" process at step 1.5 above, with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.

And while we don't undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.

Another way in which we could contrast theoretical from a-theoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification. Probably it's more complex, and probably we don't really know what a better specification would be. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are whatever those relationships are linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.

@@

WHAT CAN ML MODELS TELL US ABOUT THEORY? PRUNING VARIABLES, FUNCTIONAL FORM, INTERACTIONS, PREDICABILITY

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.  Even if B\&S were correct that in their study the theory based escalation model performed better, it would simple be a case study of one case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory.  Nor, against it.

@@

More importantly, trying to contrast machine learning and theory is misguided. Even if a "theory" section is missing, predictive models are informed by previous work and arguments. If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with cleary specified models. But we do not. What we instead have are weak theories that could be interpreted to suggest any range of equations for predicting an outcome. Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our theories.

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.  However, these comparisons must be made carefully to ensure those correct inferences are drawn.  We continue to think that theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.  Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning as well.

>>>>

---
title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  Their model is s outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory forecasting is more accurate than has previously been possible. We re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. It does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternaties over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{\cite{PITF},\cite{ICEWS},\cite{VIEWS}\cite{others}} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

1. Split the training data into training (2001 - 2007) and test (2008 - 2015) data.
2. Estimate the escalation and other competing models with the training data. 
3. Create out-of-sample (OOS) predictions from each model using the test set. 
4. Calculate AUC-ROC measures for each set of OOS predictions.

@@

The corresponding results for each question are shown in B&S Tables 1 and 2, which we examine further below. We accurately replicate their Tables 1 and 2. The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

@@

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model is not the best; it actually performs worse than the atheoretical, ktchen sink model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90%--17 of 19--clearly used empirical ROC curves, and 2 (10%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey. Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.

@@

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling [@chiba:etal:2015,@beger:etal:2017]. Although the two RFs are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger RF using both component model's underlying decision trees. Thus, while all RF models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (`ntree`), the new RF model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined RF model predicts at all, not even through a binary low-/high-risk split. The split-population PITF RF model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing RF randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

We now turn to an examination that addresses and fixes the five issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2). We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S, and may an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in *The Political Logic of Survival* [@bdm:etal:2005]. For the purposes of discussion we call this strong theory.

@@

At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are whatever those relationships are linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.

@@

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.  Even if B\&S were correct that in their study the theory based escalation model performed better, it would simply be a case study of a single case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory.  Nor, against it.

@@

More importantly, trying to contrast machine learning and theory may be misguided. Even if a "theory" section is missing, predictive models are informed by previous work and arguments. If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with cleary specified models. But we do not. What we have instead are weak theories that could be interpreted to suggest any range of equations for predicting an outcome. Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our weak theories.

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.  However, these comparisons must be made carefully to ensure those correct inferences are drawn.  We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.  Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning as well.
